SI: Settlements

### 2022-03-23 analysis/neo/neo.Rmd
add new docs edit orig paper

---
title: "SI: Settlements / Pottery groups"
author: Petr Pajdla
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
# keywords: |
#   keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

@@

**This document describes results for Pottery groups.**

@@

```{r tab-set-nr}
set_base$period2 %>%
  table_nr_set()
```

```{r map-set-nr, fig.cap = 'Settlements across time periods.'}
settlements2 %>%
  plot_nr_set()
```

@@

```{r map-kde, fig.cap = 'Kernel density estimation for different periods.'}
settlements2 %>% 
  plot_kde()
```

```{r set-kde}
kde <- settlements2 %>% 
  est_kde() %>% 
  mutate(reg = str_extract(id, "^."))
```

@@

```{r alt-data}
alt <- settlements2 %>% 
  select(id, altitude) %>% 
  st_drop_geometry() %>% 
  distinct(id, altitude) %>% 
  transmute(id, value = altitude,
            variable = "altitude")
```

```{r alt}
settlements2 %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, altitude)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Altitude (m)")
```

@@

```{r slope}
slope <- read_csv(here::here("analysis/data/derived_data", 
                             "slope_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "slope")
```

```{r plot-slope}
slope %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Slope (°)")
```

@@

```{r tpi}
tpi <- read_csv(here::here("analysis/data/derived_data", 
                           "tpi_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "tpi")
```

```{r plot-tpi}
tpi %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "TPI")
```

@@

```{r hydro}
hydro <- read_csv(here::here("analysis/data/derived_data", 
                             "hydro_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "hydro") %>% 
  rename(value = hydro_kde)
```

```{r plot-hydro}
hydro %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Hydrology KDE")
```

@@

along_lines <- along_lines$groups %>% 
  mutate(chrono = str_remove(chrono, ".$")) %>% 
  group_by(id, chrono, period, variable) %>% 
  summarise(value = max(value)) %>% 
  ungroup(id, chrono, variable) %>% 
  mutate(value = value / n()) %>% 
  select(-chrono) %>% 
  group_by(variable, period, id) %>% 
  summarise(value = mean(value), .groups = "drop")
```

@@

rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
  pivot_longer(starts_with("p"), 
               names_to = "period", values_to = "relevance") %>% 
  rename(rm = surovina)

rm_dist_rel <- set_base$period2 %>% 
  filter_rel_dist_rm()

@@

```{r variables}
res_long <- alt %>% 
  bind_rows(slope) %>% 
  bind_rows(tpi) %>% 
  bind_rows(hydro) %>% 
  left_join(set_base$period2) %>% 
  select(id, variable, value, period) %>% 
  bind_rows(along_lines) %>% 
  bind_rows(rm_dist_fin) %>% 
  bind_rows(select(kde, -reg)) %>% 
  filter(id %in% set_base$period2$id) %>% 
  group_by(variable, id, period) %>% 
  summarise(value = mean(value)) %>% 
  ungroup()

res_wide <- res_long %>% 
  pivot_wider(id_cols = c(id, period), 
              names_from = variable, 
              values_from = value, values_fill = 0)

```

@@

set_spat %>% 
  select(id) %>% 
  inner_join(classif, by = "id") %>% 
  ggplot() +
  geom_sf(aes(color = g)) +
  facet_grid(vars(period), vars(g)) +
  theme_void()
```

```{r spat-out}
# set_spat %>% 
#   select(id) %>% 
#   inner_join(classif, by = "id") %>% 
#   write_sf(here::here("analysis/groups/groups.geojson"), delete_layer = TRUE)
```

>>>>

---
title: "SI: Settlements / Neo"
author: Petr Pajdla
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
# keywords: |
#   keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

@@

**This document describes results for Neolithic (A)/B/C.**

@@

```{r}
set_base$period2$id %in% set_base$period1$id %>% table()

neoB <- c("SBK", "LgK")
neoC <- c("ENE", "TRB")

neo <- set_base$period2 %>% 
  mutate(chrono = str_remove(chrono, ".$")) %>% 
  select(-facet) %>% 
  bind_rows(set_base$period1) %>% 
  select(-starts_with("period")) %>% 
  arrange(id) %>% 
  distinct(id, chrono) %>% 
  mutate(period = if_else(chrono %in% neoB, "NeoB", NA_character_),
         period = if_else(chrono %in% neoC, "NeoC", period)) %>% 
  select(-chrono) %>% 
  distinct(id, period) %>% 
  mutate(period_label = if_else(period == "NeoB", "Neolithic B", "Neolithic C"),
         label = period_label)

neo_spat <- neo %>%
  left_join(set_spat, by = "id") %>% 
  st_as_sf()

# edit labs for chrono
labs_chrono$periods <- c("Neolithic B", "Neolithic C") %>% 
  setNames(c("NeoB", "NeoC"))
```

```{r tab-set-nr}
neo %>%
  table_nr_set()
```

```{r map-set-nr, fig.cap = 'Settlements across time periods.'}
neo_spat %>% 
  plot_nr_set()
```

@@

```{r map-kde, fig.cap = 'Kernel density estimation for different periods.'}
neo_spat %>% 
  plot_kde()
```

```{r set-kde}
kde <- neo_spat %>% 
  mutate(chrono = period) %>% 
  est_kde() %>% 
  mutate(reg = str_extract(id, "^."))
```

@@

```{r alt-data}
alt <- neo_spat %>% 
  select(id, altitude) %>% 
  st_drop_geometry() %>% 
  distinct(id, altitude) %>% 
  transmute(id, value = altitude,
            variable = "altitude")
```

```{r alt}
neo_spat %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, altitude)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Altitude (m)")
```

@@

```{r slope}
slope <- read_csv(here::here("analysis/data/derived_data", 
                             "slope_at_points.csv")) %>% 
  filter(id %in% neo$id) %>% 
  mutate(variable = "slope")
```

```{r plot-slope}
slope %>% 
  left_join(neo, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Slope (°)")
```

@@

```{r tpi}
tpi <- read_csv(here::here("analysis/data/derived_data", 
                           "tpi_at_points.csv")) %>% 
  filter(id %in% neo$id) %>% 
  mutate(variable = "tpi")
```

```{r plot-tpi}
tpi %>% 
  left_join(neo, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "TPI")
```

@@

```{r hydro}
hydro <- read_csv(here::here("analysis/data/derived_data", 
                             "hydro_at_points.csv")) %>% 
  filter(id %in% neo$id) %>% 
  mutate(variable = "hydro") %>% 
  rename(value = hydro_kde)
```

```{r plot-hydro}
hydro %>% 
  left_join(neo, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Hydrology KDE")
```

@@

along_lines <- along_lines$groups %>% 
  bind_rows(along_lines$traditions) %>% 
  mutate(chrono = str_remove(chrono, ".$")) %>% 
  select(-starts_with("period")) %>% 
  arrange(id) %>% 
  mutate(period = if_else(chrono %in% neoB, "NeoB", NA_character_),
         period = if_else(chrono %in% neoC, "NeoC", period)) %>% 
  select(-chrono) %>% 
  group_by(variable, period, id) %>% 
  summarise(value = mean(value)) %>% 
  ungroup(c(variable, period, id)) %>% 
  mutate(period_label = if_else(period == "NeoB", "Neolithic B", "Neolithic C"),
         label = period_label)
```

```{r along-lines-plot}
along_lines %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_grid(vars(variable), vars(reg)) +
  theme_bw() +
  labs(y = NULL)
```

@@

rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
  pivot_longer(starts_with("p"), 
               names_to = "period", values_to = "relevance") %>% 
  rename(rm = surovina) %>% 
  mutate(period2 = if_else(period %in% c("p5.0", "p4.8", "p4.6", "p4.4"), "NeoB", NA_character_),
         period2 = if_else(period %in% c("p4.2", "p4.0", "p3.8", "p3.6", "p3.4"), "NeoC", period2)) %>% 
  select(rm, period = period2, relevance, reg)

rm_dist_rel <- neo %>% 
  filter_rel_dist_rm()

@@

```{r variables}
res_long <- alt %>% 
  bind_rows(slope) %>% 
  bind_rows(tpi) %>% 
  bind_rows(hydro) %>% 
  left_join(neo) %>% 
  select(id, variable, value, period) %>% 
  bind_rows(select(along_lines, -ends_with("label"))) %>% 
  bind_rows(rm_dist_fin) %>% 
  bind_rows(select(kde, -reg)) %>% 
  filter(id %in% neo$id) %>% 
  group_by(variable, id, period) %>% 
  summarise(value = mean(value)) %>% 
  ungroup()

res_wide <- res_long %>% 
  pivot_wider(id_cols = c(id, period), 
              names_from = variable, 
              values_from = value, values_fill = 0)
```

@@

set_spat %>% 
  select(id) %>% 
  inner_join(classif, by = "id") %>% 
  ggplot() +
  geom_sf(aes(color = g)) +
  facet_grid(vars(g), vars(period)) +
  theme_void()
```

```{r spat-out}
# set_spat %>% 
#   select(id) %>% 
#   inner_join(classif, by = "id") %>% 
#   write_sf(here::here("analysis/neo/neo.geojson"), delete_layer = TRUE)
```

### 2022-03-23 analysis/paper/paper.Rmd
add new docs edit orig paper

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

Keywords: `r rmarkdown::metadata$keywords`

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
  mutate(rm = str_to_lower(rm))

# rm_dist_mean <- rm_dist %>%
#   select(-region, -type, -shp, -min, -median, -max) %>%
#   group_by(id, rm) %>%
#   summarise(dist_rm = mean(mean)) %>%
#   pivot_longer(cols = dist_rm, names_to = "variable") %>%
#   mutate(variable = str_c(variable, rm, sep = "_")) %>%
#   select(-rm)
```

@@

@@

```{r get-raster}
# pkg raster
# hydro_dens <- raster::raster(
#   here::here(derived_data, "hydro", "density_raster.tif"))
# 
# hydro_dens_at_points <- hydro_dens %>% 
#   raster::extract(set_spat) %>% 
#   as_tibble() %>% 
#   rename(hydro_kde = value) %>% 
#   bind_cols(select(st_drop_geometry(set_spat), id))

# pkg stars
hydro_dens <- stars::read_stars(
  here::here(derived_data, "hydro", "density_raster.tif")
) %>% 
  st_set_crs(value = st_crs(set_spat))

hydro_dens_at_points <- hydro_dens %>% 
  stars::st_extract(set_spat) %>% 
  st_drop_geometry() %>% 
  rename(hydro_kde = density_raster.tif) %>% 
  bind_cols(select(st_drop_geometry(set_spat), id))

@@

ggplot() +
  stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void() +
  labs(fill = "KDE")
```
<!--- ====================================================================== -->

@@

We focus on slope around the settlements.

```{r dem-data}
dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))

@@

alt_plot %>% 
  ggplot(aes(altitude, dem)) +
  geom_point() +
  theme_minimal() +
  annotate(geom = "text", x = 175, y = 525, 
           label = paste("Cor. ", alt_cor))
```

@@

# Settlement 'continuity'

@@

res2_long %>% 
  plot_continuity(y = n_set_per2)
```

@@

```{r variable-labels}
labs_variables <- c(
  "settlements_kde" = "Settlement density",
  "dist_fenced" = "Min. dist. to fenced set. (km)",
  "line_terrain" = "Along terrain line",
  "line_water" = "Along water line",
  "cont" = "Settlement continuity",
  "hydro_kde" = "Watercourse density",
  # "dem" = "Altitude (DEM)", 
  "altitude" = "Altitude (map)",
  "slope" = "Slope (°)",
  "rm_dist_chipped" = "Distance to chipped RM",
  "rm_dist_polished" = "Distance to polished RM"
  # "rm_pc1" = "Distance to RM sources 1",
  # "rm_pc2" = "Distance to RM sources 2",
  # "rm_pc3" = "Distance to RM sources 3"
) %>% 
  map_chr(str_wrap, 12)
```

@@

```{r plot-pca1-period1, fig.cap="PCA for pottery traditions data across periods, PC 1-2 space."}
pca1_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label")
```

```{r plot-pca1-period2, fig.cap="PCA for pottery traditions data across periods, PC 3-4 space."}
pca1_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label", c("PC3", "PC4"))
```

@@

```{r plot-pca2-period1, fig.cap="PCA for pottery groups data across periods, PC 1-2 space."}
pca2_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label")
```

```{r plot-pca2-period2, fig.cap="PCA for pottery groups data across periods, PC 3-4 space."}
pca2_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label", c("PC3", "PC4"))
```

@@

@@

mb_clust2 %>% pull(classif)

mb_clust2 %>%
    select(classif) %>% 
    ungroup(c(reg, neo)) %>%
  unnest() %>% View()
    mutate(g = str_c(reg, g))

mb_classif2_neoB <- mb_clust2 %>% mbc_classify("NeoB")
mb_classif2_neoC <- mb_clust2 %>% mbc_classify("NeoC")

>>>>

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
# keywords: |
#   keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
  mutate(rm = str_to_lower(rm))
```

@@

```{r rm-clean}
rm(list = ls(pattern = "^rm_"))
```

@@

```{r count-water-dens}
# pkg raster
# hydro_dens <- raster::raster(
#   here::here(derived_data, "hydro", "density_raster.tif"))
# 
# hydro_dens_at_points <- hydro_dens %>% 
#   raster::extract(set_spat) %>% 
#   as_tibble() %>% 
#   rename(hydro_kde = value) %>% 
#   bind_cols(select(st_drop_geometry(set_spat), id))

# pkg stars
# hydro_dens <- stars::read_stars(
#   here::here(derived_data, "hydro", "density_raster.tif")
# ) %>% 
#   st_set_crs(value = st_crs(set_spat))
# 
# hydro_dens_at_points <- hydro_dens %>% 
#   stars::st_extract(set_spat) %>% 
#   st_drop_geometry() %>% 
#   rename(hydro_kde = density_raster.tif) %>% 
#   bind_cols(select(st_drop_geometry(set_spat), id))

# pkg terra, implementing buffer around settlements
# settlements represented as points

hydro_dens <- terra::rast(
  here::here(derived_data, "hydro", "density_raster.tif")
)

hydro_dens_at_points <- terra::extract(
  hydro_dens, 
  terra::vect(st_buffer(set_spat, dist = 2e3))
) %>% 
  group_by(ID) %>% 
  summarise(hydro_kde = mean(density_raster)) %>% 
  bind_cols(id = set_spat$id) %>% 
  select(-ID)

@@

# stars
# ggplot() +
#   stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
#   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
#   theme_void() +
#   labs(fill = "KDE")
```

```{r water-clean}
rm(list = c("hydro_dens", "hydro_dens_at_points"))
```

@@

```{r dem}
dem_terra <- terra::rast(here::here(derived_data, "/dem/dem_aster_sjtsk_stars.tif"))
```

## Slope

```{r slope}
dem_slope <- terra::terrain(dem_terra, v = "slope")

# slope in the vicinity of settlements - 2km
slope <- terra::extract(dem_slope, 
                        terra::vect(st_buffer(set_spat, dist = 2e3)))

# derive median aspect
mean_slope <- slope %>% 
  tibble() %>% 
  group_by(ID) %>% 
  summarise(mean_slope = mean(slope)) %>% 
  bind_cols(id = set_spat$id) %>% 
  mutate(reg = str_extract(id, "^."))

# shapiro.test(filter(mean_slope, reg == "B")$mean_slope)
# hist(filter(mean_slope, reg == "B")$mean_slope)
# qqnorm(filter(mean_slope, reg == "B")$mean_slope)
# 
# shapiro.test(filter(mean_slope, reg == "M")$mean_slope)
# hist(filter(mean_slope, reg == "M")$mean_slope)
# qqnorm(filter(mean_slope, reg == "M")$mean_slope)

# mean_slope %>% 
#   inner_join(settlements2, by = "id") %>% 
#   ggplot(aes(mean_slope)) + 
#   geom_histogram(color = "black", fill = "white") +
#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
#   theme_bw() +
#   labs(x = "slope (°)")

# mean_slope %>% 
#   inner_join(settlements2, by = "id") %>% 
#   ggplot(aes(mean_slope, period_label)) + 
#   geom_boxplot(color = "black", fill = "white") +
#   facet_wrap(vars(reg), scales = "free_y") +
#   theme_bw() +
#   labs(x = "slope (°)")
```

```{r slope-to-res}
res1_long <- mean_slope %>% 
  select(-ID, -reg) %>% 
  inner_join(st_drop_geometry(select(settlements1, id, period))) %>% 
  transmute(id, 
            period,
            value = mean_slope, 
            variable = "slope") %>% 
  bind_rows(res1_long)

res2_long <- mean_slope %>% 
  select(-ID, -reg) %>% 
  inner_join(st_drop_geometry(select(settlements2, id, period))) %>% 
  transmute(id, 
            period,
            value = mean_slope, 
            variable = "slope") %>% 
  bind_rows(res2_long)
```

```{r aspect-clean}
rm(list = c("dem_slope", "slope", "mean_slope"))
gc()
```

<!--
```{r dem-data}
dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))

@@

alt_plot %>% 
ggplot(aes(altitude, dem)) +
geom_point() +
theme_minimal() +
annotate(geom = "text", x = 175, y = 525, 
label = paste("Cor. ", alt_cor))
```
-->

## Exploring aspect

Aspect is calculated from the DEM, 0 means the slope is facing North, 
90 East, 180 South and 270 West. The settlements are buffered 
by 200 meters and the median aspect is taken.
The vicinity of settlements is thus of interest, not the given point.

In both Morava river drainage basin and Eastern part of 
Bohemia, median aspect is normally distributed.

```{r aspect-norm}
shapiro.test(filter(med_asp, reg == "B")$med_asp)
shapiro.test(filter(med_asp, reg == "M")$med_asp)
```

```{r aspect-plot-reg, fig.cap="Histogram of aspect distibution on polar coordinates."}

# polar plot
med_asp %>% 
  ggplot(aes(med_asp)) + 
  geom_histogram(color = "black", fill = "white") +
  scale_x_continuous(limits = c(0, 359), 
                     breaks = c(0, 90, 180, 270)) +
  coord_polar(start = 0) + 
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(x = "median aspect (°)")
```

```{r aspect-plot-reg-p2}
med_asp %>% 
  inner_join(settlements2, by = "id") %>% 
  ggplot(aes(med_asp)) + 
  geom_histogram(color = "black", fill = "white") +
  scale_x_continuous(limits = c(0, 360), 
                     breaks = c(0, 90, 180, 270, 360)) +
  facet_grid(vars(period_label), vars(reg), scales = "free_y") +
  theme_bw() +
  labs(x = "median aspect (°)")
```

```{r aspect-clean}
rm(list = c("dem_asp", "asp", "med_asp"))
```

## TPI

> Terrain / Topographic roughness index

```{r tpi}
dem_tpi <- terra::terrain(dem_terra, v = "TPI")

# tpi in the vicinity of settlements - 2km
tpi <- terra::extract(dem_tpi, 
                      terra::vect(st_buffer(set_spat, dist = 2e3)))

# derive median aspect
mean_slope <- slope %>% 
  tibble() %>% 
  group_by(ID) %>% 
  summarise(mean_slope = mean(slope)) %>% 
  bind_cols(id = set_spat$id) %>% 
  mutate(reg = str_extract(id, "^."))

# shapiro.test(filter(mean_slope, reg == "B")$mean_slope)
# hist(filter(mean_slope, reg == "B")$mean_slope)
# qqnorm(filter(mean_slope, reg == "B")$mean_slope)
# 
# shapiro.test(filter(mean_slope, reg == "M")$mean_slope)
# hist(filter(mean_slope, reg == "M")$mean_slope)
# qqnorm(filter(mean_slope, reg == "M")$mean_slope)

# mean_slope %>% 
#   inner_join(settlements2, by = "id") %>% 
#   ggplot(aes(mean_slope)) + 
#   geom_histogram(color = "black", fill = "white") +
#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
#   theme_bw() +
#   labs(x = "slope (°)")

# mean_slope %>% 
#   inner_join(settlements2, by = "id") %>% 
#   ggplot(aes(mean_slope, period_label)) + 
#   geom_boxplot(color = "black", fill = "white") +
#   facet_wrap(vars(reg), scales = "free_y") +
#   theme_bw() +
#   labs(x = "slope (°)")
```

@@

<!--
# Settlement 'continuity'

@@

res2_long %>% 
plot_continuity(y = n_set_per2)
```
-->

@@

```{r variable-labels}
labs_variables <- c(
  "settlements_kde" = "Settlement density",
  "dist_fenced" = "Min. dist. to fenced set. (km)",
  "line_terrain" = "Along terrain line",
  "line_water" = "Along water line",
  # "cont" = "Settlement continuity",
  "hydro_kde" = "Watercourse density",
  # "dem" = "Altitude (DEM)", 
  "altitude" = "Altitude (map)",
  "slope" = "Slope (°)",
  "rm_dist_chipped" = "Distance to chipped RM",
  "rm_dist_polished" = "Distance to polished RM"
  # "rm_pc1" = "Distance to RM sources 1",
  # "rm_pc2" = "Distance to RM sources 2",
  # "rm_pc3" = "Distance to RM sources 3"
) %>% 
  map_chr(str_wrap, 12)
```

@@

```{r plot-pca1-period1, fig.cap="PCA for pottery traditions data across periods, PC 1-2 space."}
pca1_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label")

# ggsave(here::here("analysis/paper/temp/pca_trad_1.png"), width = 10, height = 21)
```

```{r plot-pca1-period2, fig.cap="PCA for pottery traditions data across periods, PC 3-4 space."}
pca1_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label", c("PC3", "PC4"))

# ggsave(here::here("analysis/paper/temp/pca_trad_2.png"), width = 10, height = 14)
```

@@

```{r plot-pca2-period1, fig.cap="PCA for pottery groups data across periods, PC 1-2 space."}
pca2_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label")

ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
```

```{r plot-pca2-period2, fig.cap="PCA for pottery groups data across periods, PC 3-4 space."}
pca2_per %>% 
  add_period_label() %>% 
  plot_nested_pca("period_label", c("PC3", "PC4"))

ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
```

@@

```{r package-mclust}
library(mclust)
```

@@

x <- mb_clust2 %>%
  dplyr::select(reg, neo, classif) %>% 
  tidyr::unnest() %>% 
  dplyr::ungroup(c(reg, neo)) %>% 
  dplyr::mutate(g = str_c(reg, g))

set_spat %>% 
  inner_join(x, by = "id") %>% 
  ggplot() +
  geom_sf(aes(color = g)) +
  facet_grid(vars(g), vars(neo))

mb_classif2_neoB <- mb_clust2 %>% mbc_classify(neo = "NeoB")
mb_classif2_neoC <- mb_clust2 %>% mbc_classify(neo = "NeoC")
